\section{Privileged instructions}
\label{privileged-instructions}

To ease and speed-up proving time, the zkEVM supports custom, privileged instructions that can only be executed by the kernel.
Any appearance of those privileged instructions in a contract bytecode for instance would result in an unprovable state.

In what follows, we denote by $p_{BN}$ the characteristic of the BN254 curve base field, curve for which Ethereum supports the 
ecAdd, ecMul and ecPairing precompiles.

\begin{enumerate}[align=left]
  \item[0x0C.] \texttt{ADDFP254}. Pops 2 elements from the stack interpreted as BN254 base field elements, and pushes their addition modulo $p_{BN}$ onto the stack.

  \item[0x0D.] \texttt{MULFP254}. Pops 2 elements from the stack interpreted as BN254 base field elements, and pushes their product modulo $p_{BN}$ onto the stack.

  \item[0x0E.] \texttt{SUBFP254}. Pops 2 elements from the stack interpreted as BN254 base field elements, and pushes their difference modulo $p_{BN}$ onto the stack.
  This instruction behaves similarly to the SUB (0x03) opcode, in that we subtract the second element of the stack from the initial (top) one.

  \item[0x0F.] \texttt{SUBMOD}. Pops 3 elements from the stack, and pushes the modular difference of the first two elements of the stack by the third one.
  It is similar to the SUB instruction, with an extra pop for the custom modulus.

  \item[0x21.] \texttt{KECCAK\_GENERAL}. Pops 4 elements (successively the context, segment, and offset portions of a Memory address, followed by a length $\ell$)
  and pushes the hash of the memory portion starting at the constructed address and of length $\ell$. It is similar to KECCAK256 (0x20) instruction, but can be applied to
  any memory section (i.e. even privileged ones).

  \item[0x49.] \texttt{PROVER\_INPUT}. Pushes a single prover input onto the stack.

  \item[0xC0-0xDF.] \texttt{MSTORE\_32BYTES}. Pops 4 elements from the stack (successively the context, segment, and offset portions of a Memory address, and then a value), and pushes
  a new offset' onto the stack. The value is being decomposed into bytes and written to memory, starting from the reconstructed address. The new offset being pushed is computed as the
  initial address offset + the length of the byte sequence being written to memory. Note that similarly to PUSH (0x60-0x7F) instructions there are 31 MSTORE\_32BYTES instructions, each
  corresponding to a target byte length (length 0 is ignored, for the same reasons as MLOAD\_32BYTES, see below). Writing to memory an integer fitting in $n$ bytes with a length $\ell > n$ will
  result in the integer being truncated. On the other hand, specifying a length $\ell$ greater than the byte size of the value being written will result in padding with zeroes. This
  process is heavily used when resetting memory sections (by calling MSTORE\_32BYTES\_32 with the value 0).

  \item[0xF6.] \texttt{GET\_CONTEXT}. Pushes the current context onto the stack. The kernel always has context 0.

  \item[0xF7.] \texttt{SET\_CONTEXT}. Pops the top element of the stack and updates the current context to this value. It is usually used when calling another contract or precompile,
  to distinguish the caller from the callee.

  \item[0xF8.] \texttt{MLOAD\_32BYTES}. Pops 4 elements from the stack (successively the context, segment, and offset portions of a Memory address, and then a length $\ell$), and pushes
  a value onto the stack. The pushed value corresponds to the U256 integer read from the big-endian sequence of length $\ell$ from the memory address being reconstructed. Note that an
  empty length is not valid, nor is a length greater than 32 (as a U256 consists in at most 32 bytes). Missing these conditions will result in an unverifiable proof.

  \item[0xF9.] \texttt{EXIT\_KERNEL}. Pops 1 element from the stack. This instruction is used at the end of a syscall, before proceeding to the rest of the execution logic.
  The popped element, \textit{kexit\_info}, contains several informations like the current program counter, current gas used, and if we are in kernel (i.e. privileged) mode.

  \item[0xFB.] \texttt{MLOAD\_GENERAL}. Pops 3 elements (successively the context, segment, and offset portions of a Memory address), and pushes the value stored at this memory
  address onto the stack. It can read any memory location, general (similarly to MLOAD (0x51) instruction) or privileged.

  \item[0xFC.] \texttt{MSTORE\_GENERAL}. Pops 4 elements (successively a value, then the context, segment, and offset portions of a Memory address), and writes the popped value from
  the stack at the reconstructed address. It can write to any memory location, general (similarly to MSTORE (0x52) / MSTORE8 (0x53) instructions) or privileged.

\end{enumerate}
